---
title: "Efficiency graph example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{efficiency_graph_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PFUSetup)
library(pins)
```


## Introduction

This vignette shows how to create a graph of country efficiencies first presented 
at the International Exergy Economics Workshop in 2023 (IEEW2023) 
by Matthew Kuerpus Heun (Calvin University). 
The steps involved are:

1. create the pinboard,
2. read efficiency data
3. prepare the data, including filtering, 
4. create an initial version of graph, and
5. refine the graph with colors and other details.


## Create the pinboard

The first step is to establish the pinboard for the `pins` package.
Someone from the PFUDatabase team will have shared the "PipelineReleases"
folder if you have access to IEA data. 
The `PFUSetup` package is helpful to find the correct folder.
Install the package with the following code.

```{r, eval = FALSE}
devtools::install_github("EnergyEconomyDecoupling/PFUSetup")
```

Next, create the pinboard from the correct Dropbox folder.
Your `pinboard_folder` might be different from this, 
due to a different location on your computer.

```{r}
pinboard_folder <- PFUSetup::get_abs_paths()[["pipeline_releases_folder"]]
pinboard_folder
pinboard <- pins::board_folder(pinboard_folder, versioned = TRUE)
pinboard
```


## Read efficiency data

Next, users can consult the `versions and products.xlsx` file
located in the "PipelineReleases" folder.
to find the desired database product.
For this example, 
we want data for `pf`, `fu`, and `pu` efficiencies, 
which can be found in the product whose pin is named "agg_eta_pfu".
The "versions and products.xlsx" file indicates that
the pin version of the "agg_eta_pfu" product for database v1.1 is
"20230619T051304Z-f653c".
Putting it all together, 
the following code reads the desired version of the desired data 
for v1.1 of the PFU database.

```{r}
agg_eta_pfu_data <- pins::pin_read(board = pinboard,
                                   name = "agg_eta_pfu",
                                   version = "20230619T051304Z-f653c")
```


## Prepare the data, including filtering

`agg_eta_pfu_data` is an `R` data frame with the following columns:

```{r}
colnames(agg_eta_pfu_data)
```

"Country" contains 3-letter country abbreviations.
"Method" contains only "PCM", inducating that the
"physical content method" is used for all data.
"Energy.type" contains either "E" (for energy) or "X" (for exergy).
"Year" is between 1960 and 2020, inclusive.
"IEAMW" contains one of "IEA" (IEA data), 
"MW" (muscle work), or "Both" (for both IEA and muscle work data).
"Chopped.mat" and "Chopped.var" are reserved for future use in the database.
In this object, both are always "None."
"Product.aggregation" and "Industry.aggregation" 
tell how energy carriers and industries 
are aggregated were aggregated prior to 
summing to primary, final, and useful energy stages.
For this example, primary, final, and useful aggregations should be 
the same, regardless of the value of 
"Product.aggregation" or "Industry.aggregation." 
"GrossNet" contains either "Gross" or "Net"
telling whether energy industry own use is included in 
final and useful aggregations.
"EX.p", "EX.f", and "EX.u" contain primary, final, and useful stage 
energy and exergy aggregations. 
Finally, we come to the columns of interest for this graph.
The columns names "eta_pf", "eta_fu", and "eta_pu" contain
primary-to-final, final-to-useful, and primary-to-useful efficiencies.

In the code below, we select and filter the data frame
to focus on the data of interest for this graph.
Comments in the code narrate the process.

```{r}
agg_eta_pfu_prepared <- agg_eta_pfu_data |> 
  # Select only those rows where we have both IEA and MW data, 
  # where gross energy production is provided, and
  # where aggregations were calculated from fully specified data.
  dplyr::filter(IEAMW == "Both",
                GrossNet == "Gross", 
                Product.aggregation == "Specified",
                Industry.aggregation == "Specified") |> 
  # Select only the columns of interest to us
  dplyr::select(Country, Energy.type, Year, eta_pf, eta_fu, eta_pu) |> 
  # Pivot the data to create a value column
  tidyr::pivot_longer(cols = c(eta_pf, eta_fu, eta_pu),
                      names_to = "Var.name",
                      values_to = "Value") |>
  # Make changes to the data
  dplyr::mutate(
    # We use the string "Asia_" to indicate
    # our aggregation for Asia.
    # Change to "Asia" (without the trailing "_").
    Country = dplyr::case_when(
      Country == "Asia_" ~ "Asia",
      TRUE ~ Country
    ),
    # Change "E" and "X" to "Energy" and "Exergy"
    Energy.type = dplyr::case_match(
      Energy.type,
      "E" ~ "Energy",
      "X" ~ "Exergy"
    ),
    Var.name = dplyr::case_match(
      Var.name,
      # Change names of efficiencies
      # for later subscripting of "pf", "fu", and "pu"
      "eta_pf" ~ "eta[pf]",
      "eta_fu" ~ "eta[fu]",
      "eta_pu" ~ "eta[pu]"
    ),
    # Set ordering for efficiency variables
    # so they look nice in our graph
    Var.name = factor(Var.name, levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  )
agg_eta_pfu_prepared
```
