---
title: "Efficiency graph example"
output: rmarkdown::html_vignette
# output: pdf_document
vignette: >
  %\VignetteIndexEntry{efficiency_graph_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(MKHthemes)
library(PFUSetup)
library(pins)
```


## Introduction

This vignette shows how to create a graph of country efficiencies first presented 
at the International Exergy Economics Workshop in 2023 (IEEW2023) 
by Matthew Kuerpus Heun (Calvin University). 
The steps involved are:

1. create the pinboard,
2. read efficiency data
3. prepare the data, including filtering, 
4. create an initial version of graph, and
5. refine the graph with colors and other details.


## Create the pinboard

The first step is to establish the pinboard for the `pins` package.
Someone from the PFUDatabase team will have shared the "PipelineReleases"
folder if you have access to IEA data. 
The `PFUSetup` package is helpful to find the correct folder.
Install the package with the following code.

```{r, eval = FALSE}
devtools::install_github("EnergyEconomyDecoupling/PFUSetup")
```

Next, create the pinboard from the correct Dropbox folder.
Your `pinboard_folder` might be different from this, 
due to a different location on your computer.

```{r}
pinboard_folder <- PFUSetup::get_abs_paths()[["pipeline_releases_folder"]]
pinboard_folder
pinboard <- pins::board_folder(pinboard_folder, versioned = TRUE)
pinboard
```

Later, we will use the package `MKHthemes`
to make nice-looking graphs. 
It can be loaded with the following code.

```{r, eval=FALSE}
devtools::install_github("MatthewHeun/MKHthemes")
```


## Read efficiency data

Next, users can consult the `versions and products.xlsx` file
located in the "PipelineReleases" folder.
to find the desired database product.
For this example, 
we want data for `pf`, `fu`, and `pu` efficiencies, 
which can be found in the product whose pin is named "agg_eta_pfu".
The "versions and products.xlsx" file indicates that
the pin version of the "agg_eta_pfu" product for database v1.1 is
"20230619T051304Z-f653c".
Putting it all together, 
the following code reads the desired version of the desired data 
for v1.1 of the PFU database.

```{r}
agg_eta_pfu_data <- pins::pin_read(board = pinboard,
                                   name = "agg_eta_pfu",
                                   version = "20230619T051304Z-f653c")
```


## Prepare the data, including filtering

`agg_eta_pfu_data` is an `R` data frame as shown below:

```{r}
agg_eta_pfu_data
```

The columns have the following meanings:

* "Country" contains 3-letter country abbreviations.
* "Method" contains only "PCM", inducating that the
"physical content method" is used for all data.
* "Energy.type" contains either "E" (for energy) or "X" (for exergy).
* "Year" is between 1960 and 2020, inclusive.
* "IEAMW" contains one of "IEA" (IEA data), 
* "MW" (muscle work), or "Both" (for both IEA and muscle work data).
* "Chopped.mat" and "Chopped.var" are reserved for future use in the database.
In this object, both are always "None."
* "Product.aggregation" and "Industry.aggregation" 
tell how energy carriers and industries 
are aggregated were aggregated prior to 
summing to primary, final, and useful energy stages.
For this example, primary, final, and useful aggregations should be 
the same, regardless of the value of 
* "Product.aggregation" or "Industry.aggregation." 
* "GrossNet" contains either "Gross" or "Net"
telling whether energy industry own use is included in 
final and useful aggregations.
* "EX.p", "EX.f", and "EX.u" contain primary, final, and useful stage 
energy and exergy aggregations. 
* Finally, we come to the columns of interest for this graph.
The columns names "eta_pf", "eta_fu", and "eta_pu" contain
primary-to-final, final-to-useful, and primary-to-useful efficiencies.

In the code below, we select and filter the data frame
to focus on the data of interest for this graph.
Comments in the code narrate the process.

```{r}
agg_eta_pfu_prepared <- agg_eta_pfu_data |> 
  # Select only those rows where we have both IEA and MW data, 
  # where gross energy production is provided, and
  # where aggregations were calculated from fully specified data.
  dplyr::filter(IEAMW == "Both",
                GrossNet == "Gross", 
                Product.aggregation == "Specified",
                Industry.aggregation == "Specified") |> 
  # Select only the columns of interest to us
  dplyr::select(Country, Energy.type, Year, eta_pf, eta_fu, eta_pu) |> 
  # Pivot the data to create a value column
  tidyr::pivot_longer(cols = c(eta_pf, eta_fu, eta_pu),
                      names_to = "Var.name",
                      values_to = "Value") |>
  # Make changes to the data
  dplyr::mutate(
    # We use the string "Asia_" to indicate
    # our aggregation for Asia.
    # Change to "Asia" (without the trailing "_").
    Country = dplyr::case_when(
      Country == "Asia_" ~ "Asia",
      TRUE ~ Country
    ),
    # Change "E" and "X" to "Energy" and "Exergy"
    Energy.type = dplyr::case_match(
      Energy.type,
      "E" ~ "Energy",
      "X" ~ "Exergy"
    ),
    Var.name = dplyr::case_match(
      Var.name,
      # Change names of efficiencies
      # for later subscripting of "pf", "fu", and "pu"
      "eta_pf" ~ "eta[pf]",
      "eta_fu" ~ "eta[fu]",
      "eta_pu" ~ "eta[pu]"
    ),
    # Set ordering for efficiency variables
    # so they look nice in our graph
    Var.name = factor(Var.name, levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  )
# Here is the prepared data
agg_eta_pfu_prepared
```



## Create an initial version of graph

At this point, 
`agg_eta_pfu_prepared` is a data frame that contains
everything we need to make the graph.
We use the `ggplot2` package to make the graph.

```{r, fig.retina=4, fig.align='center', fig.height=4, fig.width=6}
agg_eta_pfu_prepared |> 
  # Set the x and y axes
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year,
                                         y = Value,
                                         # Group by Country 
                                         # to make individual lines
                                         group = Country)) +
  # Set a fine line width
  ggplot2::geom_line(linewidth = 0.05) +
  # Set the years to appear on the x axis
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  # Set labels on each axis
  ggplot2::labs(x = NULL,
                y = expression(eta)) +
  # Separate data into mini-graphs by 
  # energy type (rows) and efficiency (columns)
  ggplot2::facet_grid(rows = ggplot2::vars(Energy.type),
                      cols = ggplot2::vars(Var.name),
                      # label_parsed gives 
                      # greek letters and subscripts
                      labeller = ggplot2::label_parsed) +
  # Make a nice gray border, no gridlines, inward-pointing ticks, 
  # and attractive spacing between axis labels and the graphs
  MKHthemes::xy_theme()
```



## Refine the graph with colors and other details

The graph contains all the data we want but has no pizzazz.
To improve its appearance,
we can to colour the individual
country lines by continent.
And, we can highlight the world as a thick line 
atop all countries.

The first step is to join continent information to the 
efficiency data, thereby making continent
data available for the graph.
We can load a data frame of countries and continents
and join it ot the efficiency data frame.

```{r}
# The continents file is an input to the pipeline that creates the database.
# Users can create their own continent mapping
# or ask for ours.
continents_file <- PFUSetup::get_abs_paths()[["aggregation_mapping_path"]]
continents_file
continents_agg_map <- continents_file |>
  # Read the Excel file
  readxl::read_excel(sheet = "continent_aggregation") |>
  # Adjust some of the column names
  dplyr::rename(
    Country = Many,
    Continent = Few
  ) |>
  dplyr::mutate(
    # Fix "Asia_" --> "Asia"
    # to match the efficiency data frame.
    Continent = dplyr::case_when(
      Continent == "Asia_" ~ "Asia",
      TRUE ~ Continent
    )
  )
continents_agg_map
# Join to the efficiency data
agg_eta_pfu_with_continents <- agg_eta_pfu_prepared |>
  dplyr::left_join(continents_agg_map, by = "Country")
agg_eta_pfu_with_continents
```

The following code creates another version of the graph,
this time with colors for continents.

First, establish the colors and graphical details we want.

```{r}
continent_colours <- c(Africa = "orange", Asia = "gray",
                       Bunkers = "darkgray", Europe = "darkgreen",
                       MidEast = "cadetblue3",
                       NoAmr = "red",
                       Oceania = "blue", SoCeAmr = "violet",
                       World = "black")
```

Now make the graph.
Details different from the previous version are
identified by comments.

```{r, fig.retina=4, fig.align='center', fig.height=4, fig.width=6}
# Create a data frame of only World information
# for years >- 1971 
# (the first year when all countries are in IEA data)
agg_eta_pfu_World <- agg_eta_pfu_with_continents |> 
  dplyr::filter(Country == "World", Year >= 1971)
# Create a data frame of countries only
agg_eta_pfu_countries_only <- agg_eta_pfu_with_continents |> 
  dplyr::filter(!(Country %in% c("Africa", "Asia", "Bunkers", "Europe", 
                                 "MidEast", "NoAmr", "Oceania", "SoCeAmr", 
                                 "World", "WRLD")))

ggplot2::ggplot(mapping = ggplot2::aes(x = Year,
                                       y = Value,
                                       group = Country, 
                                       # Set country colours by continent
                                       colour = Continent)) +
  # Create two different line layers.
  # The first has countries only.
  ggplot2::geom_line(data = agg_eta_pfu_countries_only,
                     linewidth = 0.05) +
  # The second (upper) layer has World only.
  ggplot2::geom_line(data = agg_eta_pfu_World, 
                     linewidth = 1, 
                     colour = "black") +
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  # Use the continent_colours for the lines
  ggplot2::scale_colour_manual(values = continent_colours) +
  ggplot2::labs(x = NULL,
                y = expression(eta), 
                # Eliminate the colour label, "Continent".
                # We all know these are continents.
                colour = NULL) +
  ggplot2::facet_grid(rows = ggplot2::vars(Energy.type),
                      cols = ggplot2::vars(Var.name),
                      labeller = ggplot2::label_parsed) +
  MKHthemes::xy_theme() +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      # Adjust the size of the line segments in the legend
      # independent from linewidths in the graph itself
      override.aes = list(linewidth = 1)
    )
  )
```


## Conclusion

The PFU Database provides data to make graphs 
such as primary-to-final, 
final-to-useful, and
primary-to-useful efficiencies.
This vignette shows details of 
how to read and clean database information, 
how to make such graphs, and 
how to prettify the graphs.
