---
title: "Efficiency graph example"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{efficiency_graph_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  # Set several aspects of graph chunks
  fig.retina=4,
  fig.align='center', 
  fig.height=4,
  fig.width=6
)
```


## Introduction

The MR-PFU Database contains primary, final, and useful 
energy and exergy data for all countries covered 
by the International Energy Agency (IEA).
Aggregations are also available. 

But it can be a challenge to get started with the MR-PFU Database
if you are unfamiliar with some types of `R` programming. 
This vignette provides a detailed example of working with PFU Database information
to help users get over that hurdle.

In particular, 
this vignette shows how to create a graph of country 
energy and exergy efficiencies first presented 
at the International Exergy Economics Workshop in 2023 (IEEW2023) 
by Matthew Kuperus Heun (Calvin University).
In the process, 
important concepts for accessing and working with the data 
are demonstrated, 
including 

1. creating a pinboard,
2. reading efficiency data,
3. preparing the data, including filtering, 
4. creating an initial version of graph, and
5. refining the graph for presentation.

Most `R` programs use packages that are loaded at the beginning
of a session.
The following packages are used in this vignette. 

```{r setup, message=FALSE}
library(devtools)
library(dplyr)
library(ggplot2)
library(MKHthemes)
library(PFUSetup)
library(pins)
library(readxl)
library(tidyr)
```

All packages are available on the 
Comprehensive R Archive Network (CRAN),
except `MKHthemes` and `PFUSetup`. 
Instructions for loading 
`MKHthemes` and `PFUSetup`
are given below.


## Create the pinboard

The first step is to establish a "pinboard" for the `pins` package.
Someone from the PFU Database team will have shared the "PipelineReleases"
folder if you have access to IEA data. 
The `PFUSetup` package is helpful to find the correct folder.
Install the `PFUSetup` package with the following code.

```{r, eval = FALSE}
devtools::install_github("EnergyEconomyDecoupling/PFUSetup")
```

Create the pinboard from the correct Dropbox folder.
Your `pinboard_folder` might be different from this, 
due to a different location on your computer.

```{r}
pinboard_folder <- PFUSetup::get_abs_paths()[["pipeline_releases_folder"]]
pinboard_folder
pinboard <- pins::board_folder(pinboard_folder, versioned = TRUE)
pinboard
```


## Read efficiency data

Next, users can consult the `versions and products.xlsx` file
located in the "PipelineReleases" folder.
to find the desired database product.
For this example, 
we want data for `pf`, `fu`, and `pu` efficiencies, 
which can be found in the product whose pin is named "agg_eta_pfu".
The "versions and products.xlsx" file indicates that
the pin version of the "agg_eta_pfu" product for database v1.1 is
"20230619T051304Z-f653c".
Putting it all together, 
the following code reads the desired version of the desired data 
for v1.1 of the PFU database.

```{r}
agg_eta_pfu_data <- pins::pin_read(board = pinboard,
                                   name = "agg_eta_pfu",
                                   version = "20230619T051304Z-f653c")
```


## Prepare the data, including filtering

`agg_eta_pfu_data` is an `R` data frame as shown below:

```{r}
agg_eta_pfu_data
```

The columns have the following meanings:

* "Country" contains 3-letter country abbreviations.
* "Method" contains only "PCM", inducating that the
"physical content method" is used for all data.
* "Energy.type" contains either "E" (for energy) or "X" (for exergy).
* "Year" is between 1960 and 2020, inclusive.
* "IEAMW" contains one of "IEA" (IEA data), 
* "MW" (muscle work), or "Both" (for both IEA and muscle work data).
* "Chopped.mat" and "Chopped.var" are reserved for future use in the database.
In this object, both are always "None."
* "Product.aggregation" and "Industry.aggregation" 
tell how energy carriers and industries 
are aggregated were aggregated prior to 
summing to primary, final, and useful energy stages.
For this example, primary, final, and useful aggregations should be 
the same, regardless of the value of 
* "Product.aggregation" or "Industry.aggregation." 
* "GrossNet" contains either "Gross" or "Net"
telling whether energy industry own use is included in 
final and useful aggregations.
* "EX.p", "EX.f", and "EX.u" contain primary, final, and useful stage 
energy and exergy aggregations. 
* Finally, we come to the columns of interest for this graph.
The columns names "eta_pf", "eta_fu", and "eta_pu" contain
primary-to-final, final-to-useful, and primary-to-useful efficiencies.

In the code below, we prepare the data 
by selecting and filtering the `agg_eta_pfu_data` data frame
to focus on the data of interest for this graph.
Comments in the code narrate the process.

```{r}
agg_eta_pfu_prepared <- agg_eta_pfu_data |> 
  # Select only those rows where we have both IEA and MW data, 
  # where gross energy production is provided, and
  # where aggregations were calculated from fully specified data.
  dplyr::filter(IEAMW == "Both",
                GrossNet == "Gross", 
                Product.aggregation == "Specified",
                Industry.aggregation == "Specified") |> 
  # Select only the columns of interest to us
  dplyr::select(Country, Energy.type, Year, eta_pf, eta_fu, eta_pu) |> 
  # Pivot the data to create a value column
  tidyr::pivot_longer(cols = c(eta_pf, eta_fu, eta_pu),
                      names_to = "Var.name",
                      values_to = "Value")
# Here is the prepared data
agg_eta_pfu_prepared
```



## Create an initial version of graph

At this point, 
`agg_eta_pfu_prepared` is a data frame that contains
everything we need to make the graph.
We use the `ggplot2` package to make the graph.

```{r}
agg_eta_pfu_prepared |> 
  # Set the x and y axes
  ggplot2::ggplot(mapping = ggplot2::aes(x = Year,
                                         y = Value,
                                         # Group by Country 
                                         # to make individual lines
                                         group = Country)) +
  # Show the data with lines
  ggplot2::geom_line() +
  # Put mini-graphs for energy type in rows and efficiency type in columns
  ggplot2::facet_grid(rows = ggplot2::vars(Energy.type),
                      cols = ggplot2::vars(Var.name))
```



## Refine the graph with colors and other details

The initial version of the graph provides confidence that we can 
use the data, but it is not attractive.

* Line widths should be reduced so we can see 
  individual countries instead of an undifferentiated 
  mass of lines for all countries.
* Lines could be coloured by continent. 
* A world average line should be added and distinguished
  from other lines.
* The column labels ("eta_pf", "eta_fu", and "eta_pu") 
  should be changed to Greek letters with subscripts
  ($\eta_{pf}$, $\eta_{fu}$, and $\eta_{pu}$).
* The left-to-right order of efficiencies should be changed to
  "eta_pf", "eta_fu", "eta_pu".
* "E" and "X" on the rows of the graph should be changed to
  "Energy" and "Exergy" for clarity.
* Change "Asia_" to "Asia" in the "Country" and "Continent" column
  (to be added later).
* The word "Value" on the $y$-axis should be changed to 
  they symbol for efficiency ($\eta$).
* There is no need for the word "Year" on the $x$-axis. 
  Everyone knows those are years.
* The year labels run into each other.
* The gray background is ugly.

Many of these changes are made by adjusting the 
values in the metadata columns 
of the `agg_eta_pfu_prepared` data frame.
The code below shows those adjustments.

```{r}
agg_eta_pfu_nice <- agg_eta_pfu_prepared |> 
  dplyr::mutate(
    # We use the string "Asia_" to indicate
    # our aggregation for Asia.
    # Change to "Asia" (without the trailing "_").
    Country = dplyr::case_when(
      Country == "Asia_" ~ "Asia",
      TRUE ~ Country
    ),
    # Change "E" and "X" to "Energy" and "Exergy"
    Energy.type = dplyr::case_match(
      Energy.type,
      "E" ~ "Energy",
      "X" ~ "Exergy"
    ),
    Var.name = dplyr::case_match(
      Var.name,
      # Change names of efficiencies
      # for later subscripts of "pf", "fu", and "pu"
      "eta_pf" ~ "eta[pf]",
      "eta_fu" ~ "eta[fu]",
      "eta_pu" ~ "eta[pu]"
    ),
    # Set ordering for efficiency variables
    # so they look nice in our graph
    Var.name = factor(Var.name, levels = c("eta[pf]", "eta[fu]", "eta[pu]"))
  )
```

To further improve appearance,
we can colour the individual
country lines by continent.
And, we can highlight the world as a thick line 
atop all countries.

The first step is to join continent information to the 
efficiency data, thereby making continent
data available for the graph.
We can load a data frame of countries and continents
and join it to the efficiency data frame.

```{r}
# The continents file is an input 
# to the calculation pipeline 
# that created the database.
# Users can create their own 
# continent mapping data frame
# or ask for ours.
continents_file <- PFUSetup::get_abs_paths()[["aggregation_mapping_path"]]
continents_file
continents_agg_map <- continents_file |>
  # Read the continent_aggregation tab of the Excel file
  readxl::read_excel(sheet = "continent_aggregation") |>
  # Adjust the column names
  dplyr::rename(
    Country = Many,
    Continent = Few
  ) |>
  dplyr::mutate(
    # Fix "Asia_" --> "Asia"
    # to match the efficiency data frame.
    Continent = dplyr::case_when(
      Continent == "Asia_" ~ "Asia",
      TRUE ~ Continent
    )
  )
continents_agg_map
# Join to the efficiency data
agg_eta_pfu_with_continents <- agg_eta_pfu_nice |>
  dplyr::left_join(continents_agg_map, by = "Country")
# Notice the Continent column
agg_eta_pfu_with_continents
```

Establish the continent colors.

```{r}
continent_colours <- c(Africa = "orange", Asia = "gray",
                       Bunkers = "darkgray", Europe = "darkgreen",
                       MidEast = "cadetblue3",
                       NoAmr = "red",
                       Oceania = "blue", SoCeAmr = "violet",
                       World = "black")
```

To create a specific World-average line, 
we need to separate the World data from 
all countries.
Also, continent averages should be removed
from the data frame that contains country lines.

```{r}
# Create a data frame of only World information
# for years >- 1971 
# (the first year when all countries are in IEA data).
# This data frame will be used to create the 
# World-average line
agg_eta_pfu_world <- agg_eta_pfu_with_continents |> 
  dplyr::filter(Country == "World", Year >= 1971)
# Create a data frame of countries only, 
# eliminating continent and world averages
agg_eta_pfu_countries <- agg_eta_pfu_with_continents |> 
  dplyr::filter(!(Country %in% c("Africa", "Asia", "Bunkers", "Europe", 
                                 "MidEast", "NoAmr", "Oceania", "SoCeAmr", 
                                 "World", "WRLD")))
```

We will use the package `MKHthemes`
to make nice-looking graphs with a white background
and tick marks that point inward, not outward. 
`MKHthemes` can be loaded with the following code.

```{r, eval=FALSE}
devtools::install_github("MatthewHeun/MKHthemes")
```


Now make the graph.
Details different from the previous version are
identified by comments.

```{r}
ggplot2::ggplot(mapping = ggplot2::aes(x = Year,
                                       y = Value,
                                       group = Country, 
                                       # Set country colours by continent
                                       colour = Continent)) +
  # Create two different line layers.
  # The first has countries only and a narrow linewidth.
  ggplot2::geom_line(data = agg_eta_pfu_countries,
                     linewidth = 0.05) +
  # The second (upper) layer has World only, 
  # a larger linewidth and black colour.
  ggplot2::geom_line(data = agg_eta_pfu_world, 
                     linewidth = 1, 
                     colour = "black") +
  # Set only a few years on the x-axis so the labels
  # don't collide
  ggplot2::scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  # Use the continent_colours for the lines
  ggplot2::scale_colour_manual(values = continent_colours) +
  # Eliminate the x-axis label ("Year"); 
  # everyone knows those are years.
  # Eliminate the colour label, "Continent";
  # everyone knows these are continent names.
  ggplot2::labs(x = NULL,
                y = expression(eta), 
                colour = NULL) +
  ggplot2::facet_grid(rows = ggplot2::vars(Energy.type),
                      cols = ggplot2::vars(Var.name),
                      labeller = ggplot2::label_parsed) +
  # Set pleasant aesthetics such as white background, 
  # gray border lines, and 
  # inward-pointing tick marks.
  MKHthemes::xy_theme() +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      # Adjust the size of the line segments in the legend
      # independent from linewidths in the graph itself
      override.aes = list(linewidth = 1)
    )
  )
```

The new version of the graph 
is aesthetically pleasing and conveys more information
than the initial attempt.


## Conclusion

The PFU Database provides data to make graphs 
such as primary-to-final, 
final-to-useful, and
primary-to-useful efficiencies.
This vignette shows details of 
how to read and clean data, 
how to make such graphs, and 
how to make the graphs more aesthetically pleasing.
